Follow-up Samples

### 2017-08-14 submission/manuscript_R1.tex
move figs and tables to after references

\textbf{Results.} There were small changes to the bacterial community
associated with adenoma or advanced adenoma and large changes associated
with carcinoma. The communities from patients with carcinomas changed
significantly more than those with adenoma following treatment (P-value
\textless{} 0.001). Although treatment was associated with intrapersonal
changes, the change in the abundance of individual OTUs in response to
treatment was not consistent within diagnosis groups (P-value
\textgreater{} 0.05). Because the distribution of OTUs across patients
and diagnosis groups was irregular, we used the Random Forest machine
learning algorithm to identify groups of OTUs that could be used to
classify pre and post-treatment samples for each of the diagnosis
groups. Although the three models could differentiate between the pre
and post-treatment samples, there was little overlap between the OTUs
that were indicative of each treatment. To determine whether individuals
who underwent treatment were more likely to have OTUs associated with
normal colons we used a larger cohort that contained individuals with
normal colons and those with adenomas, advanced adenomas, and
carcinomas. We again built Random Forest models and measured the change
in the positive probability of having one of the three diagnoses to
assess whether the post-treatment samples received the same
classification as the pre-treament samples. Samples from patients who
had carcinomas changed towards a microbial milieu that resembles the
normal colon after treatment (P-value \textless{} 0.05). Finally, we
were unable to detect any significant differences in the microbiota of
individuals treated with surgery alone and those treated with
chemotherapy or chemotherapy and radiation (P-value \textgreater{}
0.05).

\textbf{Conclusions.} By better understanding the response of the
microbiota to treatment for adenomas and carcinomas, it is likely that
biomarkers will eventually be validated that can be used to quantify the
risk of recurrence and the likelihood of survival. Although it was
difficult to identify significant differences between pre and
post-treatment samples from patientes with adenoma and advanced adenoma,
this was not the case for carcinomas. Not only were there large changes
in pre versus post-treatment samples for those with carcinoma, but these
changes were towards a more normal microbiota.

@@

We answered these questions by sequencing the V4 region of 16S rRNA
genes amplified from fecal samples of individuals with adenoma, advanced
adenoma, and carcinomas pre and post-treatment. We used classical
community analysis to compare the alpha and beta-diversity of
communities pre and post-treatment. Next, we generated Random Forest
models to identify bacterial populations that were indicative of
treatment for each diagnosis group. Finally, we measured the predictive
probabilities to assess whether treatment yielded bacterial communities
similar to those individuals with normal colons. We found that treatment
alters the composition of the gut microbiota and that, for those with
carinomas, the gut microbiota shifted more towards that of a normal
colon after treatment. In the individuals with carcinomas, no difference
was found by the type of treatment (surgery alone, surgery with
chemotherapy, surgery with chemotherapy and radiation). Understanding
how the community responds to these treatments could be a valuable tool
for identifying biomarkers to quantify the risk of recurrence and the
likelihood of survival.

@@

\textbf{\emph{Treatment for colonic lesions alters the bacterial
community structure}} Within our 67-person cohort we tested whether the
microbiota of patients with adenoma (N = 22), advanced adenoma (N = 19),
or carcinoma (N = 26) had any broad differences between pre and
post-treatment samples {[}Table 1{]}. None of the individuals in this
study had any recorded antibiotic usage that was not associated with
surgical treatment of their respective lesion. The structure of the
microbial communities of the pre and post-treatment samples differed, as
measured by the \(\theta\)\textsubscript{YC} beta diversity metric
{[}Figure 1A{]}. We found that the communities obtained pre and
post-treament among the patients with carcinomas changed significantly
more than those patients with adenoma (P-value \textless{} 0.001). There
were no significant differences in the amount of change observed between
the patients with adenoma and advanced adenoma or between the patients
with advanced adenoma and carcinoma (P-value \textgreater{} 0.05). Next,
we tested whether there was a consistent direction in the change in the
community structure between the pre and post-treatment samples for each
of the diagnosis groups {[}Figure 1B-D{]}. We only observed a consistent
shift in community structure for the patients with carcinoma when using
a PERMANOVA test (adenoma P-value = 0.999, advanced adenoma P-value =
0.945, and carcinoma P-value = 0.005). Finally, we measured the number
of observed OTUs, Shannon evenness, and Shannon diversity in the pre and
post-treatment samples and did not observe a significant change for any
of the diagnosis groups (P-value \textgreater{} 0.05) {[}Table S1{]}.

\textbf{\emph{The treatment of lesions are not consistent across
diagnosis groups.}} We used two approaches to identify those bacterial
populations that change between the two samples for each diagnosis
group. First, we sought to identify individual OTUs that could account
for the change in overall community structure. However, using a paired
Wilcoxon test we were unable to identify any OTUs that were
significantly different in the pre and post-treatment groups (P-value
\textgreater{} 0.05). It is likely that high inter-individual variation
and the irregular distribution of OTUs across individuals limited the
statistical power of the test. We attempted to overcome these problems
by using Random Forest models to identify collections of OTUs that would
allow us to differentiate between pre and post-treatment samples from
each of the diagnosis groups. To limit the likelihood that the models
would overfit the data because of the relatively small number of
subjects in each group, we restricted our models to only incorporate 10
OTUs. Despite this restriction, the models performed well (adenoma AUC
range = 0.69 - 0.92, advanced adenoma AUC range = 0.80 - 1.00, carcinoma
AUC range = 0.82 - 0.98). Interestingly, the 10 OTUs that were used for
each model had little overlap with each other {[}Figure 2{]}. Although
treatment had an impact on the overall community structure, the effect
of treatment was not consistent across patients and diagnosis groups.

@@

This study expands upon existing research that has established a role
for the microbiota in tumorigenesis and that demonstrated the utility of
microbiome-based biomarkers to predict the presence of colonic lesions.
We were surprised by the lack of a consistent signal that was associated
with treatment of patients with adenomas or advanced adenomas. The lack
of a large effect size may be due to differences in the role of bacteria
in the formation of adnomas and carcinomas or it could be due to
differences in the behaviors and medications within these classes of
patients. One of the most exciting of these future directions is the
possibility that markers within the microbiota could be used to
potentially evaluate the effect of treatment and predict recurrence for
those diagnosed with carcinoma. If such an approach is effective, it
might be possible to target the microbiota as part of adjuvant therapy,
if the biomarkers identified play a key role in the disease process. Our
data provides additional evidence on the importance of the microbiota in
tumorigenesis by addressing the recovery of the microbiota after
treatment and opens interesting avenues of research into how these
changes may affect recurrence.

@@

\textbf{\emph{Treatment.}} For this study treatment refers specifically
to the removal of a lesion with or without chemotherapy and radiation.
The majority of patients undergoing treatment for adenoma or advanced
adenoma were not treated surigically {[}Table 1{]} but rather via
colonoscopy. All patients diagnosed with carcinomas were treated with at
least surgery or a combination of surgery and chemotherapy or surgery,
chermotherapy, and radiation. The type of chemotherapy used for patients
with CRC included Oxaliplatin, Levicovorin, Folfox, Xeloda,
Capecitabine, Avastin, Fluorouracil, and Glucovorin. These were used
individually or in combination with others depending on the patient
{[}Table 1{]}. If an individual was treated with radiation they were
also always treated with chemotherapy. Radiation therapy generally used
18 mV photons for treatment.

@@

Similarily, the Random Forest {[}33{]} algorithm was also used to create
the three models used to classify normal versus diagnosis. These samples
were obtained using the same methodology as described earlier in this
section. All samples used for this component of model training were from
pre-treatment samples only. The total number of individuals in the
normal versus diagnosis models was 423 individuals {[}Table 2{]}. There
were a total of 239 individuals in the normal versus adenoma model, 262
individuals in the normal versus advanced adenoma model, and 266
individuals in the normal versus carcinoma model {[}Table 2{]}.

@@

For each of the pre versus post-treamtent models assessment of the most
important OTUs was then made by taking the top 10 OTUs by mean decrease
in accuracy (MDA). These were then used to build each respective reduced
OTU pre versus post-treatment model by diagnosis group to help avoid
model overfitting. These reduced models were then put through the same
process mentioned in the previous paragraph and were what was used for
the final pre versus post-treatment models. For the normal versus
diagnosis models the important OTUs were obtained by counting the number
of times an OTU was present in the top 10\% of MDA for each of the 100
different splits. This was then followed with filtering of this list to
variables that were only present in more than 50\% of these 100 runs.
These corresponding reduced OTU normal versus diagnosis models were then
put through the same process mentioned in the previous paragraph and
were what was used for the final normal versus diagnosis models. For the
pre versus post-treatment models the final optimized mtry was 2 and for
the normal versus diagnosis models the final optimized mtry was 2.

Each model was then applied to our 67-person cohort {[}Table 1{]} based
on diagnosis: adenoma (pre-treatment adenoma (adenoma n = 22 and disease
free n = 0) versus post-treatment adenoma (adenoma n = 0 and disease
free n = 22)), advanced adenoma pre-treatment advanced adenoma (advanced
adenoma n = 19 and disease free n = 0 ) versus post-treatment advanced
adenoma (advanced adenoma n = 0 and disease free n = 19), and carcinoma
(pre-treatment carcinoma (carinoma n = 26 and disease free n = 0) versus
post-treatment carcinoma (carcinoma n = 1 and disease free n = 25)). The
application of the pre versus post-treatment models generated the
probabilites that the sample was a pre-treatment sample. The application
of the normal versus diagnosis models generated the probabilities that
the sample was that specific diagnosis (adenoma, advanced adenoma, or
carcinoma).

@@

\subsection*{References}\label{references}
\addcontentsline{toc}{subsection}{References}

>>>>

\textbf{Results.} There were small changes to the bacterial community
associated with adenoma or advanced adenoma and large changes associated
with carcinoma. The communities from patients with carcinomas changed
significantly more than those with adenoma following treatment (P-value
\textless{} 0.001). Although treatment was associated with intrapersonal
changes, the change in the abundance of individual OTUs in response to
treatment was not consistent within diagnosis groups (P-value
\textgreater{} 0.05). Because the distribution of OTUs across patients
and diagnosis groups was irregular, we used the Random Forest machine
learning algorithm to identify groups of OTUs that could be used to
classify pre and post-treatment samples for each of the diagnosis
groups. Although the three models could differentiate between the pre
and post-treatment samples, there was little overlap between the OTUs
that were indicative of each treatment. To determine whether individuals
who underwent treatment were more likely to have OTUs associated with
normal colons we used a larger cohort that contained individuals with
normal colons and those with adenomas, advanced adenomas, and
carcinomas. We again built Random Forest models and measured the change
in the positive probability of having one of the three diagnoses to
assess whether the post-treatment samples received the same
classification as the pre-treatment samples. Samples from patients who
had carcinomas changed towards a microbial milieu that resembles the
normal colon after treatment (P-value \textless{} 0.05). Finally, we
were unable to detect any significant differences in the microbiota of
individuals treated with surgery alone and those treated with
chemotherapy or chemotherapy and radiation (P-value \textgreater{}
0.05).

\textbf{Conclusions.} By better understanding the response of the
microbiota to treatment for adenomas and carcinomas, it is likely that
biomarkers will eventually be validated that can be used to quantify the
risk of recurrence and the likelihood of survival. Although it was
difficult to identify significant differences between pre and
post-treatment samples from patients with adenoma and advanced adenoma,
this was not the case for carcinomas. Not only were there large changes
in pre versus post-treatment samples for those with carcinoma, but these
changes were towards a more normal microbiota.

@@

We answered these questions by sequencing the V4 region of 16S rRNA
genes amplified from fecal samples of individuals with adenoma, advanced
adenoma, and carcinomas pre and post-treatment. We used classical
community analysis to compare the alpha and beta-diversity of
communities pre and post-treatment. Next, we generated Random Forest
models to identify bacterial populations that were indicative of
treatment for each diagnosis group. Finally, we measured the predictive
probabilities to assess whether treatment yielded bacterial communities
similar to those individuals with normal colons. We found that treatment
alters the composition of the gut microbiota and that, for those with
carcinomas, the gut microbiota shifted more towards that of a normal
colon after treatment. In the individuals with carcinomas, no difference
was found by the type of treatment (surgery alone, surgery with
chemotherapy, surgery with chemotherapy and radiation). Understanding
how the community responds to these treatments could be a valuable tool
for identifying biomarkers to quantify the risk of recurrence and the
likelihood of survival.

@@

\textbf{\emph{Treatment for colonic lesions alters the bacterial
community structure}} Within our 67-person cohort we tested whether the
microbiota of patients with adenoma (N = 22), advanced adenoma (N = 19),
or carcinoma (N = 26) had any broad differences between pre and
post-treatment samples {[}Table 1{]}. None of the individuals in this
study had any recorded antibiotic usage that was not associated with
surgical treatment of their respective lesion. The structure of the
microbial communities of the pre and post-treatment samples differed, as
measured by the \(\theta\)\textsubscript{YC} beta diversity metric
{[}Figure 1A{]}. We found that the communities obtained pre and
post-treatment among the patients with carcinomas changed significantly
more than those patients with adenoma (P-value \textless{} 0.001). There
were no significant differences in the amount of change observed between
the patients with adenoma and advanced adenoma or between the patients
with advanced adenoma and carcinoma (P-value \textgreater{} 0.05). Next,
we tested whether there was a consistent direction in the change in the
community structure between the pre and post-treatment samples for each
of the diagnosis groups {[}Figure 1B-D{]}. We only observed a consistent
shift in community structure for the patients with carcinoma when using
a PERMANOVA test (adenoma P-value = 0.999, advanced adenoma P-value =
0.945, and carcinoma P-value = 0.005). Finally, we measured the number
of observed OTUs, Shannon evenness, and Shannon diversity in the pre and
post-treatment samples and did not observe a significant change for any
of the diagnosis groups (P-value \textgreater{} 0.05) {[}Table S1{]}.

\textbf{\emph{The treatment of lesions are not consistent across
diagnosis groups.}} We used two approaches to identify those bacterial
populations that change between the two samples for each diagnosis
group. First, we sought to identify individual OTUs that could account
for the change in overall community structure. However, using a paired
Wilcoxon test we were unable to identify any OTUs that were
significantly different in the pre and post-treatment groups (P-value
\textgreater{} 0.05). It is likely that high inter-individual variation
and the irregular distribution of OTUs across individuals limited the
statistical power of the test. We attempted to overcome these problems
by using Random Forest models to identify collections of OTUs that would
allow us to differentiate between pre and post-treatment samples from
each of the diagnosis groups. To limit the likelihood that the models
would over-fit the data because of the relatively small number of
subjects in each group, we restricted our models to only incorporate 10
OTUs. Despite this restriction, the models performed well (adenoma AUC
range = 0.69 - 0.92, advanced adenoma AUC range = 0.80 - 1.00, carcinoma
AUC range = 0.82 - 0.98). Interestingly, the 10 OTUs that were used for
each model had little overlap with each other {[}Figure 2{]}. Although
treatment had an impact on the overall community structure, the effect
of treatment was not consistent across patients and diagnosis groups.

@@

This study expands upon existing research that has established a role
for the microbiota in tumorigenesis and that demonstrated the utility of
microbiome-based biomarkers to predict the presence of colonic lesions.
We were surprised by the lack of a consistent signal that was associated
with treatment of patients with adenomas or advanced adenomas. The lack
of a large effect size may be due to differences in the role of bacteria
in the formation of adenomas and carcinomas or it could be due to
differences in the behaviors and medications within these classes of
patients. One of the most exciting of these future directions is the
possibility that markers within the microbiota could be used to
potentially evaluate the effect of treatment and predict recurrence for
those diagnosed with carcinoma. If such an approach is effective, it
might be possible to target the microbiota as part of adjuvant therapy,
if the biomarkers identified play a key role in the disease process. Our
data provides additional evidence on the importance of the microbiota in
tumorigenesis by addressing the recovery of the microbiota after
treatment and opens interesting avenues of research into how these
changes may affect recurrence.

@@

\textbf{\emph{Treatment.}} For this study treatment refers specifically
to the removal of a lesion with or without chemotherapy and radiation.
The majority of patients undergoing treatment for adenoma or advanced
adenoma were not treated surgically {[}Table 1{]} but rather via
colonoscopy. All patients diagnosed with carcinomas were treated with at
least surgery or a combination of surgery and chemotherapy or surgery,
chemotherapy, and radiation. The type of chemotherapy used for patients
with CRC included Oxaliplatin, Levicovorin, Folfox, Xeloda,
Capecitabine, Avastin, Fluorouracil, and Glucovorin. These were used
individually or in combination with others depending on the patient
{[}Table 1{]}. If an individual was treated with radiation they were
also always treated with chemotherapy. Radiation therapy generally used
18 mV photons for treatment.

@@

Similarly, the Random Forest {[}33{]} algorithm was also used to create
the three models used to classify normal versus diagnosis. These samples
were obtained using the same methodology as described earlier in this
section. All samples used for this component of model training were from
pre-treatment samples only. The total number of individuals in the
normal versus diagnosis models was 423 individuals {[}Table 2{]}. There
were a total of 239 individuals in the normal versus adenoma model, 262
individuals in the normal versus advanced adenoma model, and 266
individuals in the normal versus carcinoma model {[}Table 2{]}.

@@

For each of the pre versus post-treatment models assessment of the most
important OTUs was then made by taking the top 10 OTUs by mean decrease
in accuracy (MDA). These were then used to build each respective reduced
OTU pre versus post-treatment model by diagnosis group to help avoid
model over-fitting. These reduced models were then put through the same
process mentioned in the previous paragraph and were what was used for
the final pre versus post-treatment models. For the normal versus
diagnosis models the important OTUs were obtained by counting the number
of times an OTU was present in the top 10\% of MDA for each of the 100
different splits. This was then followed with filtering of this list to
variables that were only present in more than 50\% of these 100 runs.
These corresponding reduced OTU normal versus diagnosis models were then
put through the same process mentioned in the previous paragraph and
were what was used for the final normal versus diagnosis models. For the
pre versus post-treatment models the final optimized mtry was 2 and for
the normal versus diagnosis models the final optimized mtry was 2.

Each model was then applied to our 67-person cohort {[}Table 1{]} based
on diagnosis: adenoma (pre-treatment adenoma (adenoma n = 22 and disease
free n = 0) versus post-treatment adenoma (adenoma n = 0 and disease
free n = 22)), advanced adenoma pre-treatment advanced adenoma (advanced
adenoma n = 19 and disease free n = 0 ) versus post-treatment advanced
adenoma (advanced adenoma n = 0 and disease free n = 19), and carcinoma
(pre-treatment carcinoma (carcinoma n = 26 and disease free n = 0)
versus post-treatment carcinoma (carcinoma n = 1 and disease free n =
25)). The application of the pre versus post-treatment models generated
the probabilities that the sample was a pre-treatment sample. The
application of the normal versus diagnosis models generated the
probabilities that the sample was that specific diagnosis (adenoma,
advanced adenoma, or carcinoma).

@@

\subsection{References}\label{references}
